---
title: "Forecasting: Principles and practice"
subtitle: "Chapters 2-3: Time series graphics and decomposition"
format: 
    pdf:
        number-sections: true 
        toc: true
execute: 
  echo: true
  output: false
engine: knitr
---

```{r initiate-reticulate}
library('easypackages')
packages = c('reticulate', 'fpp3', 'tidyverse', 'fable')
libraries(packages)

use_condaenv("dtp_shiny_v20", required = TRUE)
```

```{python loading-packages}
import os
import tarfile
import urllib
import pandas as pd
import numpy as np
from plotnine import * 

os.chdir('notebooks/')
```

```{python load-data-pyt}
df_gas = pd.read_csv('../data/raw/fueltypesall.csv')

# convert date to datetime object 
df_gas['Date'] = pd.to_datetime(df_gas['Date'])
```


```{r load-data-r}
df_gasr = read_csv('../data/raw/fueltypesall.csv')

# convert date to tsibble and then to long format
df_gasr <- df_gasr |>
  pivot_longer(cols = -c(Date, `Fuel Type`, `Type de carburant`), 
               names_to = "city", values_to = "price") |> 
  mutate(week= yearweek(Date)) |>
  as_tsibble(index = week, key = c('city', `Fuel Type`))
```


## Time plots

```{r time-plots}
# plot gas prices in toronto over time
toronto = df_gasr |>
  filter(city == 'Toronto East/Est')

plot_toronto = autoplot(toronto) + 
  labs(y='price') + 
  theme_classic()

ggsave("figures/toronto_plot.png", plot = plot_toronto, width = 8, height = 6)
```

```{python time-plots-py}
# Create the same plot using Python 
plot_toronto = (
    ggplot(df_gas, aes(x='Date', y='Toronto East/Est', color ="Fuel Type")) +
    geom_line() +
    labs(y='price') +
    theme_classic()
)
```

 
## Time series patterns 

1. **Trend**: a long-term change that does not have to be linear. 
2. **Seasonal**: changes that occur for fixed and known periods (e.g., holidays, seasons)
3. **Cyclic**: rises and falls that are not of a fixed frequency (e.g., economic conditions) and last at least 2 years. 

Plots below examine quarterly data. 

```{r r-quarterly}
aus_production |> 
  filter(year(Quarter) >= 1980) |>
  autoplot(Electricity) +
  labs(y='gWh', title='Australian electricity production') + 
  theme_classic()
```

```{python python-quarterly}
(ggplot(r.aus_production, aes(x='Quarter', y='Electricity')) +
    geom_line() +
    labs(y='price') +
    theme_classic()
)
```


# Time series decomposition


Adjustments are often useful as they create simpler time series. There are four common types of adjustments: 

1) Calendar: Instead of computing average sales per month, reduce it down to sales/day, as the number of selling days differ by month. 
2) Population: Similarly, compute population variables at to either per capita or relative to some fixed denominator (1000)
3) Inflation: adjust financial data such that they are expressed in common dollars (i.e., adjusted for infation). 
4) Mathematical: if data are heteroscedastic, log transformations are useful (and also still interpretable). 

## Time series components

There are two types of decompositions: 

1. **Additive**. Value at any given $y_t$ time point is the sum of a seasonal ($S_t$), trend-cycle ($T_t$), and remainder ($R_t$) component. 

$$
y_t = S_t + T_t + R_t
$$

2. **Mulitplicative**. Value at any given $y_t$ time point is the product of a seasonal ($S_t$), trend-cycle ($T_t$), and remainder ($R_t$) component. 

$$
y_t = S_t \times T_t \times  R_t
$$

Additive decompoisiton is appropriate if neither $S_t$ or $T_t$ vary with the level of $y_t$. If not, multiplicative decomposition is more appropriate. Alternative, if there is heteroscedasity in either the seasonal or trend-cycle components, then transformations can be applied to instill stability in the data. On this point, note that using a log transformation is equivalent to using the multiplicative decomposition. 

### Examle: Employment in the US retail sector

```{r retail-plot}

us_retail_employment = us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)

autoplot(us_retail_employment, Employed) +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```




